Constraint Programming - Inside Gecode

### 2014-05-03 paper.tex
constraint propagation

	\subsection{Constraint Types}
	As already mentioned before constraint programming is the generalization of linear programming, therefore it is at least possible do define linear constraints like \refeq{eq:linearConstraint}. Which statements are possible to make varies than very much from the programming language in which the problem is implemented and the used toolset. Typical constraints would be
	\begin{itemize}
		\item Equalities: $x = y + 2$, $x^2 + y^2 = 1$, \dots
		\item Inequalities: $x \ne y$, \dots
		\item Relations: $x < 2 y$, $e^x < 4$, \dots
	\end{itemize}
	In Gecode for example it is very easy to create for example a nonlinear constraint based on the integer variables a, b, c and d \cite[p.~120]{programmingGecode}:
	\begin{lstlisting}
rel(home, a+b*(c+d) == 0);
	\end{lstlisting}
	This single statement in C++ implements the constraint
	\begin{equation}
		a + b \cdot (c + d) == 0
	\end{equation}
	which is a quite handy way to do it.

	\begin{figure}
	\center
	\begin{tikzpicture}
		\draw (0, 0) rectangle (1, 1);
		\node at (0, -0.3) {($x_1$, $y_1$)};
		\draw[arrows=<->] (1.2, 0) -- (1.2, 1);
		\node at (1.5, 0.5) {$d_1$};
		\draw (1.5, 1.5) rectangle(2, 2);
		\node at (1.5, 1.2) {($x_2$, $y_2$)};
		\draw[arrows=<->] (2.2, 1.5) -- (2.2, 2);
		\node at (2.5, 1.75) {$d_2$};
	\end{tikzpicture}
	\caption{two squares which should not overlap}
	\label{fig:squares}
	\end{figure}
	Hence there are boolean variables available it is also possible to use logical operators to combine other constraint types to logical expressions. To show the usage of this on an example I will give the implementation for the placement of two squares such that they do not overlap \reffig{fig:squares}. This problem can be expressed mathematically by \cite[p. 101]{programmingGecode}
	\begin{equation}
	\begin{split}
		x_1 + d_1 \le x_2 & \lor x_2 + d_2 \le x_1 \lor \\
		y_1 + d_1 \le y_2 & \lor y_2 + d_2 \le y_1
	\end{split}
	\end{equation}
	and in Gecode by
	\begin{lstlisting}
rel(home,	(x1 + d1 <= x2) || 
		(x2 + d2 <= x1) || 
		(y1 + d1 <= y2) || 
		(y2 + d2 <= y1));
	\end{lstlisting}
	which is again quite obvious to a reader with some experience in a C-like programming language.

@@

	To visualize this effect consider a constraint problem based on two variables $x \in \{1, 5, 6\}$ and $y \in \{1, 2\}$ and one constraint based on them.
	\begin{equation}
		c_1:\ x \le y
	\end{equation}
	Without constraint propagation and branching first on $x$ this results in search tree \reffig{fig:bigTree}. Obviously the right parts of these tree can already be pruned at the root with propagation of constraint $c_1$ and therefore it is not necessary to create and evaluate these subtrees \reffig{fig:smallTree}.

@@

	Constraints in Gecode are always evaluated ahead of a further growth of a subtree, as it may be possible to prune some of the subtrees. The problem here is to decide in which order the constraints should be considered, and it may even be necessary to evaluate some constraints more often. A short example will explain this very well \cite[p.~575]{handbookCP}: Let there be two constraints 
	\begin{equation}
		c_1: y - x^2 = 0
	\end{equation}
	\begin{equation}
		c_2: y - x - 1 = 0
	\end{equation}
	and let the variable domains be $x, y \in [-4, 4]$. Starting with the evaluation of constraint one follows a chain of reductions of the variable domains, based only on constraint propagation:
	\begin{equation}
	\begin{split}
		c_1:\ &y \in [-4, 4] \rightarrow y \in [0, 4] \\
		c_1:\ &x \in [-4, 4] \rightarrow x \in [-2, 2] \\
		c_2:\ &y \in [0, 4] \rightarrow y \in [0, 3] \\
		c_2:\ &x \in [-2, 2] \rightarrow x \in [-1, 2] \\
		c_1:\ &x \in [-1, 2] \rightarrow x \in [-1, 1.73 \dots] \\
		&\dots
	\end{split}
	\end{equation}
	Actually this behaviour is even very nice to have as it narrows the down the search space very efficient.

@@

	The next difficult decision is which constraint should evaluated first. In the example above the constraints were quite simple and therefore it won't make a difference to evaluate first $c_1$ or $c_2$, the search will take nearly the same time. Unfortunately not all constraints are that easy to evaluate and for this case Gecode sets for every constraint a certain cost \cite[p.~275]{programmingGecode}. Based on this cost the constraints are sorted into buckets which are then iterated from the constraints with lower costs to the ones with higher costs. These means, whenever a variable domain is changed the first try is to use constraints with the lowest possible cost and only if none of them needs to be evaluated constraints with higher costs are considered.

>>>>

	\subsection{Constraint Types}

	\begin{figure}
	\center
	\begin{tikzpicture}
		\draw (0, 0) rectangle (1, 1);
		\node at (0, -0.3) {($x_1$, $y_1$)};
		\draw[arrows=<->] (1.2, 0) -- (1.2, 1);
		\node at (1.5, 0.5) {$d_1$};
		\draw (1.5, 1.5) rectangle(2, 2);
		\node at (1.5, 1.2) {($x_2$, $y_2$)};
		\draw[arrows=<->] (2.2, 1.5) -- (2.2, 2);
		\node at (2.5, 1.75) {$d_2$};
	\end{tikzpicture}
	\caption{two squares which should not overlap}
	\label{fig:squares}
	\end{figure}

	As already mentioned before constraint programming is the generalization of linear programming, therefore it is at least possible do define linear constraints like \refeq{eq:linearConstraint}. Which statements are possible to make varies than very much from the programming language in which the problem is implemented and the used toolset. Typical constraints would be
	\begin{itemize}
		\item Equalities: $x = y + 2$, $x^2 + y^2 = 1$, \dots
		\item Inequalities: $x \ne y$, \dots
		\item Relations: $x < 2 y$, $e^x < 4$, \dots
	\end{itemize}
	In Gecode for example it is very easy to create for example a nonlinear constraint based on the integer variables a, b, c and d \cite[p.~120]{programmingGecode}:
	\begin{lstlisting}
rel(home, a+b*(c+d) == 0);
	\end{lstlisting}
	This single statement in C++ implements the constraint
	\begin{equation}
		a + b \cdot (c + d) == 0
	\end{equation}
	which is a quite handy way to do it.

	Hence there are boolean variables available it is also possible to use logical operators to combine other constraint types to logical expressions. To show the usage of this on an example I will give the implementation for the placement of two squares such that they do not overlap \reffig{fig:squares}. This problem can be expressed mathematically by \cite[p. 101]{programmingGecode}
	\begin{equation}
	\begin{split}
		x_1 + d_1 \le x_2 & \lor x_2 + d_2 \le x_1 \lor \\
		y_1 + d_1 \le y_2 & \lor y_2 + d_2 \le y_1
	\end{split}
	\end{equation}
	and in Gecode by
	\begin{lstlisting}
rel(home,	(x1 + d1 <= x2) || 
		(x2 + d2 <= x1) || 
		(y1 + d1 <= y2) || 
		(y2 + d2 <= y1));
	\end{lstlisting}
	which is again quite obvious to a reader with some experience in a C-like programming language.

@@

	To visualize this effect consider a constraint problem based on two variables $x \in \{1, 5, 6\}$ and $y \in \{1, 2\}$ and one constraint based on them.
	\begin{equation}
		c_1:\ x \le y
	\end{equation}
	Without constraint propagation and branching first on $x$ this results in search tree \reffig{fig:bigTree}. Obviously the right parts of this tree can already be pruned already at the root by propagation of constraint $c_1$ and therefore it is not necessary to create and evaluate these subtrees \reffig{fig:smallTree}.

@@

	Constraints in Gecode are always evaluated ahead of a further growth of a subtree, as it may be possible to prune some of the subtrees. The problem here is to decide in which order the constraints should be considered, and it may even be necessary to evaluate some constraints more often. A short example will explain this very well \cite[p.~575]{handbookCP}: Let there be two constraints 
	\begin{equation}
		c_1: y - x^2 = 0
	\end{equation}
	\begin{equation}
		c_2: y - x - 1 = 0
	\end{equation}
	and let the variable domains be $x, y \in [-4, 4]$. Starting with the evaluation of constraint one follows a chain of reductions of the variable domains, based only on constraint propagation:
	\begin{equation}
	\begin{split}
		c_1:\ &y \in [-4, 4] \rightarrow y \in [0, 4] \\
		c_1:\ &x \in [-4, 4] \rightarrow x \in [-2, 2] \\
		c_2:\ &y \in [0, 4] \rightarrow y \in [0, 3] \\
		c_2:\ &x \in [-2, 2] \rightarrow x \in [-1, 2] \\
		c_1:\ &x \in [-1, 2] \rightarrow x \in [-1, 1.73 \dots] \\
		&\dots
	\end{split}
	\end{equation}
	Actually this behaviour is very nice to have as it narrows the down the search space efficiently.

@@

	The next difficult decision is which constraint should evaluated first. In the example above the constraints were quite simple and therefore it won't make a difference to evaluate first $c_1$ or $c_2$, the search will take nearly the same time. Unfortunately not all constraints are that easy to evaluate and for this case Gecode sets for every constraint a certain cost \cite[p.~275]{programmingGecode}. Based on this cost the constraints are sorted into buckets which are then iterated from the constraints with lower costs to the ones with higher costs. This means, whenever a variable domain is changed, the first try is to use constraints with the lowest possible cost and only if none of them needs to be evaluated constraints with higher costs are considered.
